Network Origins of Social Distinction

### 2025-07-27 manuscript.tex
Updates from Overleaf

\subsubsection{Distinction in Triadic Closure and Formation}

@@

In addition to these explanations, the distinction measure also reveals the utility of a distinction based theory of triadic closure. Figure 3 \textit{Model (a)} and Table 4 present a graph and corresponding measures of near-neighbor triadic closure. The distinction of nodes 1 and 6 increases to .106, compared to 0 in the circle graph, while all other nodes fall below 0 due to the reverberating effects of the new connection. Notably, node 7 increases in eigenvector status from the combination, but overall decrease in scaled distinction due to the tie. While both eigenvector centrality and scaled distinction would predict tie formation in this simplified social structure, the scaled distinction measure would find that Node 7 would actually lose out in some ways from this arrangement. This matches with findings that initial connections do weaken when triadic closure occurs \citep{huang2018will}, consistent with a reaction to the increased constraint placed on the closure broker. This example demonstrates the fragility of the circle graph as an equality condition in a distinction-based theoretical framework, as there are opportunities for mutual increases in distinction through triadic closure. Beyond questions of balancing valence as is the current framework for understanding triadic closure, a distinction-based theoretical framework can explain the propensity towards triadic closure, but also its failure to form depending on the initial patterning of relationships.

Figure 3 \textit{Model (b)} and Table 5 present a graph and corresponding measures of a long tie formation spanning the circle graph. The connected nodes 4 and 7 both have distinction maximizing incentives to form a tie, but notably their scaled distinction increases more when they close triadically with a near neighbor. Thus, under sparse network conditions, the gains in status from local consolidation outweigh the distinction gains from increased autonomy.

%actually creates hierarchy, you only one tie to create 4 levels of hierarchy
%the pyramid is the natural form of organization
%agents form ties with maximization, then you grow transitivty 
%two conditions, inital local ties matter more, with density, long ties matter more

%Circles as their own, pizza pie. at least get to 0 status. Wheel can emerge
%folds vs circles

%kite graph-outsider breaks mechanical solidarity
%mechanical solidarity is really an outcome of the network structure (Durkheim kind of %said that but Omar said it better). Try to get inequality in an organic solidarity system but that's unstable? (might have written that down wrong). 
%mechanical solidarity benefits from the stranger. more theoretical than empirical. 
%actually anthropologists ruined everything in the villages
%also somewhat the Marco Polo vibe, Shogun (the violation of markers of distinction, %didn't matter), The Last Samurai
%theory of subgraph cliques rather than full graph as clique

%gets close to 0 negative, with non-normalized eigenvectors
%if you normalize then it all works as expected
% in the normalized kite graph, 1.0, the other get close, decreases the scalar, evs don't mean anything

%college caves, BYU vs Notre Dame vs regulars

@@

    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/circle-plus-triangle.png}
            \caption{Circle Graph with Closure}
            \label{fig:circleplus}
    \end{subfigure}
    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/circle-plus-long.png}
            \caption{Circle Graph with Long Tie}
            \label{fig:circleplus}
    \end{subfigure}
    \captionsetup[subfigure]{font=footnotesize,labelfont=footnotesize}
    \centering
    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/kite.png}
            \caption{Kite Graph}
            \label{fig:kite}
    \end{subfigure}
    \caption{Distinction Centrality in toy graphs. Nodes with the highest distinction are shown in red, nodes with positive distinction are shown in blue, nodes with negative distinction are shown in tan, and nodes with distinction close to zero are shown in purple.}
    \label{fig:toys}
\end{figure}

@@

@@

\textit{Model (b)} presents an idealized depiction of a structural fold \citep{vedres2010structural}. The resulting measurement scores are in Table 2. While node 1 maintains high eigenvector centrality, the eigenvector centrality of the other nodes increases, while their constraint decreases. It is actually the constraint of Node 1 that increases, reaching 1.00 because all its neighbors are maximally interconnected in their respective subgroups. This position of brokerage is more constrained than the star graph because the other actors are connected in subgroups. The actors in the subgroup are able to combine their interests to mitigate against the independence of Node 1 in the center. In an unadjusted measure of distinction, the different nodes become remarkably close in their distinction measure. With a scalar adjustment, the stratification increases, giving Node 1 a slight advantage, but also highlighting the extent to which Node 1's distinction is eroded. In a maximally connected graph, the distinction of each node would also go to 0.00, as there is no network differentiation among nodes.

@@

The insight that it is the middling levels that are most likely to form long ties to create their distinction indicates that it is structures with a level of stratification and the opportunity to form long ties that are most likely to create cultural heterogeneity, or in other words a flourishing of subcultures. Colleges are not merely opportunities for people from many backgrounds to recombine their cultural repertoires, but are actually network structures that encourage the pursuit of cultural distinction given the long ties that accompany the students new position. The expansion of higher education beyond the old elites in the 1950s and 60s was a network transformation that beget the major cultural shifts that began in these institutions. These cultural transformations were neither inherent in the students nor the institutions, but came out of the network structure that enabled the pursuit of cultural distinction to match the newly acquired network position.

@@

%College as inherently distinction based institution makes liberals a thing, rather than if you only draw from caves

>>>>

\subsubsection{Distinction in Triadic Closure}

@@

In addition to these explanations, the distinction measure also reveals the utility of a distinction based theory of triadic closure. Figure 3 \textit{Model (a)} and Table 4 present a graph and corresponding measures of near-neighbor triadic closure. The distinction of nodes 1 and 6 increases to .106, compared to 0 in the circle graph, while all other nodes fall below 0 due to the reverberating effects of the new connection. Notably, node 7 increases in eigenvector status from the combination, but overall decrease in scaled distinction due to the tie. While both eigenvector centrality and scaled distinction would predict tie formation in this simplified social structure, the scaled distinction measure would find that Node 7 would actually lose out in some ways from this arrangement. This matches with findings that initial connections do weaken when triadic closure occurs \citep{huang2018will}, consistent with a reaction to the increased constraint placed on the closure broker. This example demonstrates the fragility of the circle graph as an equality condition in a distinction-based theoretical framework, as there are opportunities for mutual increases in distinction through triadic closure. Additionally, the  example of triadic closure shows that only mild additions to network ties can very quickly create stratification. The addition of a single connection between node 1 and 6 in Figure 3 \textit{Model (a)} stratified a network from equality into four levels of hierarchy with a single additional connection.

Beyond questions of balancing valence as is the current framework for understanding triadic closure, a distinction-based theoretical framework can explain the propensity towards triadic closure, but also its failure to form depending on the initial patterning of relationships.

Figure 3 \textit{Model (b)} and Table 5 present a graph and corresponding measures of a long tie formation spanning the circle graph. The connected nodes 4 and 7 both have distinction maximizing incentives to form a tie, but notably their scaled distinction increases more when they close triadically with a near neighbor. Thus, under sparse overall network conditions, the gains in status from local consolidation outweigh the distinction gains from increased autonomy. Conversely, once dense local conditions prevail, long ties can become more valuable as the gain in autonomy outweighs the marginal gains in status. However, the unique topography of networks requires careful consideration in each case.

What the difference between these two examples suggests is that, based on considerations for distinction, the initial spread of social organization occurs locally through the triadic balance of near neighbors. This would suggest that the natural form of organizational formation is the pyramid formed from combinations of local triads that are further connected by brokers or welds creating structural folds. However, the transitivity created by the initial triad formation would spread quickly throughout the network, making welding structural folds a highly advantageous strategy for members of triads who could facilitate that connection, which is further explored below.

@@

    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/circle-plus-triangle.png}
            \caption{Circle Graph with Near Neighbor Closure}
            \label{fig:circleplus}
    \end{subfigure}
    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/circle-plus-long.png}
            \caption{Circle Graph with Long Tie}
            \label{fig:circleplus}
    \end{subfigure}
    \caption{Distinction Centrality in toy graphs. Nodes with the highest distinction are shown in red, nodes with positive distinction are shown in blue, nodes with negative distinction are shown in tan, and nodes with distinction close to zero are shown in purple.}
    \label{fig:toys}
\end{figure}

@@

\subsubsection{Kite Graphs and Eigenvector Normalization}

%gets close to 0 negative, with non-normalized eigenvectors
%if you normalize then it all works as expected
% in the normalized kite graph, 1.0, the other get close, decreases the scalar, evs don't mean anything
%computtional experiment where you make kite less dense, or size of kite
%kite graph-outsider breaks mechanical solidarity
%mechanical solidarity is really an outcome of the network structure (Durkheim kind of %said that but Omar said it better). Try to get inequality in an organic solidarity system but that's unstable? (might have written that down wrong). 
%mechanical solidarity benefits from the stranger. more theoretical than empirical. 
%actually anthropologists ruined everything in the villages
%also somewhat the Marco Polo vibe, Shogun (the violation of markers of distinction, %didn't matter), The Last Samurai
%theory of subgraph cliques rather than full graph as clique
%hypothesis: size of the normalization gaps is causing the differences between normalized and unnormalized.

\begin{figure}
    \captionsetup[subfigure]{font=footnotesize,labelfont=footnotesize}
    \centering
    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=1.0\textwidth]{Plots/Toys/kite.png}
            \caption{Kite Graph}
            \label{fig:kite}
    \end{subfigure}
    \caption{Distinction Centrality in toy graphs. Nodes with the highest distinction are shown in red, nodes with positive distinction are shown in blue, nodes with negative distinction are shown in tan, and nodes with distinction close to zero are shown in purple.}
    \label{fig:toys}
\end{figure}

@@

%welds martin2009social
%Structural holes vs folds (need to graph out the holes)

%Circles as their own, pizza pie. at least get to 0 status. Wheel can emerge
%folds vs circles
\textit{Model (b)} presents an idealized depiction of a structural fold \citep{vedres2010structural}. The resulting measurement scores are in Table 2. While node 1 maintains high eigenvector centrality, the eigenvector centrality of the other nodes increases, while their constraint decreases. It is actually the constraint of Node 1 that increases, reaching 1.00 because all its neighbors are maximally interconnected in their respective subgroups. This position of brokerage is more constrained than the star graph because the other actors are connected in subgroups. The actors in the subgroup are able to combine their interests to mitigate against the independence of Node 1 in the center. In an unadjusted measure of distinction, the different nodes become remarkably close in their distinction measure. With a scalar adjustment, the stratification increases, giving Node 1 a slight advantage, but also highlighting the extent to which Node 1's distinction is eroded. In a maximally connected graph, the distinction of each node would also go to 0.00, as there is no network differentiation among nodes.

@@

The insight that it is the middling levels that are most likely to form long ties to create their distinction indicates that it is structures with a level of stratification and the opportunity to form long ties that are most likely to create cultural heterogeneity, or in other words a flourishing of subcultures. Colleges are not merely opportunities for people from many backgrounds to recombine their cultural repertoires, but are actually network structures that encourage the pursuit of cultural distinction given the long ties that accompany the students new position. The expansion of higher education beyond the old elites in the 1950s and 60s was a network transformation that beget the major cultural shifts that began in these institutions. These cultural transformations were neither inherent in the students nor the institutions, but came out of the network structure that enabled the pursuit of cultural distinction to match the newly acquired network position. Thus, if "liberals" are those who espouse diversity and innovation in cultural forms, to the extent that universities draw from a variety of social bases, liberal dispositions will results. However, if the university draws from an isolated basis, as the common educational destination of a delimited community, then the structural opportunities for cultural distinction are absent, and the culture is inherited from the isolated community.

@@
